Calibration

### 2023-01-02 Paper.Rmd
Results: start of calibration

---
title: "Disturbance Assimilation"
author: "Michael Dietze"
date: '`r Sys.Date()`'
output:
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    keep_tex: true
bibliography: bibliography.bibtex
---

@@

# Outline

# Abstract

@@

>>>>

---
title: "Assimilating disturbance: Toward real-time monitoring and forecasting"
author:
  - Michael C. Dietze, Earth and Environment, Boston University, Boston, MA
  - Shawn P. Serbin, Environmental and Climate Sciences Department, Brookhaven National Laboratory, Upton, NY
  - Robert Kennedy, Geography, Environmental Sciences, and Marine Resource Management, Oregon State University, Corvallis, OR
  - Beverly E. Law, Forest Ecosystems & Society, Oregon State University, Corvallis, OR
date: '`r Sys.Date()`'
output:
  pdf_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    keep_tex: true
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    keep_tex: true
bibliography: bibliography.bibtex
---

@@

# Abstract {-}

Current approaches to bottom-up disturbance monitoring rely heavily on the detection of land-use, land-use change, and forestry (LULUCF) through remote sensing, but often account for ecosystem impacts using simple look-up tables. By contrast, process models are frequently used to analyze and predict disturbance dynamics in greater detail. Once observations are available, however, we need to update predictions, especially for stochastic processes such as disturbance. State data assimilation (SDA) is designed specifically to update predictions, nudging modeled states back toward reality in proportion to the uncertainties in the model and the data, but current SDA algorithms are designed to update continuous states, not discrete disturbances. Here we develop a new Bayesian SDA algorithm that combines a discrete Multinomial state-and-transition framework with conventional ensemble filtering SDA approaches. To demonstrate the potential for assimilating disturbance, we applied the Multinomial SDA to the Very Simple Ecosystem Model (VSEM), performing both simulated data experiments with known disturbances and testing the algorithm against real-world disturbances detected in the LandTrender data product for central Oregon.

With simulated disturbance we demonstrate the ability to not only detect discrete disturbance events but also avoid false positives. We also demonstrated the ability to fuse multiple data types to successfully distinguish different disturbance types, and to probabilistically capture vegetation type ‘switching’ events within the assimilation and ensemble forecast. To apply this to real-world data we calibrated VSEM against eddy-covariance and ancillary data from the Ameriflux US-Me2 tower. We then selected 356 conifer forest sites for testing, using the Landtrendr disturbance product to stratify by four disturbance types (cut, burn, pest, and other). We then assimilated the 30m Landtrendr annual aboveground biomass product from 1990-2017 and assessed the rate of disturbance detection. Assimilating just AGB, our assimilation was sensitive to disturbances that reduced biomass by 1.5 kg/m2 but underpredicted defoliation disturbances, which we expect would be improved by also assimilating LAI. Moving forward, the SDA framework provides an exciting opportunity to fuse multiple data sources to holistically improve real-time disturbance detection, impact assessment (e.g. carbon sequestration), and forecasts of both disturbance events and post-disturbance recovery within a single integrated system.

@@

<!---- SUPPLEMENTAL MATERIAL --->
```{r child = 'supplemental.Rmd'}
```

### 2023-01-02 calibrate.Rmd
Results: simulated data experiments
Results: start of calibration

AF.biomass = raster::extract(bioS,AF[,c("Lat.","Long.")])

@@

@@

save(theta,Ql,Qw,Qs,file="fit.RData")
```

>>>>

bioS = stack("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
AF.biomass = raster::extract(bioS,AF[,c("Lat.","Long.")])

@@

## calibration only
Cw.post = tapply(post[,"CR"],years,mean)
plot(year,Cw.post,ylab="AGB (kgC/m2)",type='l',lwd=3,
     xlim=c(2002,2012),ylim=range(Cw,na.rm=TRUE))
points(year,Cw,col=2,pch="+",cex=2)
legend("topleft",legend=c("model","data"),pch=c(NA,"+"),lty=c(1,NA),col=1:2)

@@

## R2
RMSE = sqrt(c(mean((Cl.post-Cl)^2,na.rm=TRUE),
            mean((Cw.post-Cw)^2,na.rm=TRUE),
            mean((NEEm-aNEE)^2,na.rm=TRUE)))
1-RMSE^2/c(var(Cl,na.rm = TRUE),var(Cw,na.rm = TRUE),var(aNEE))

save(theta,Ql,Qw,Qs,NEEm,aNEE,year,Cl,Cl.post,Cw,Cw.post,mle,file="fit.RData")
```

### 2023-01-02 methods.Rmd
Results: start of calibration

```{r concept, echo=FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

## assimilation
priors <- list(p = p.dist,
                 mufN = mu1,
                 mufD = mu2,
                 pfN  = 1/P1,
                 pfD  = 1/P2,
                 Y = obs,
                 r = 1/obs.sd^2
                 )
mod <- jags.model(file=textConnection(AnalysisSSSD),
                   data=priors,n.adapt=1000,n.chains=3)
jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)

>>>>

```{r concept, echo=FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

## assimilation
AnalysisSSSD <- "
model{
  ## prior
  muN ~ dnorm(mufN,pfN)
  muD ~ dnorm(mufD,pfD)
  D ~ dbern(p)
  ## Analysis
  n <- D*muD + (1-D)*muN
  Y ~ dnorm(n,r)
}"
priors <- list(p = p.dist,
                 mufN = mu1,
                 mufD = mu2,
                 pfN  = 1/P1,
                 pfD  = 1/P2,
                 Y = obs,
                 r = 1/obs.sd^2
                 )
mod <- jags.model(file=textConnection(AnalysisSSSD),
                   data=priors,n.adapt=1000,n.chains=3)
jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)

### 2023-01-02 results.Rmd
Results: start of calibration

>>>>

Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.  Best fit parameters are given in Supplemental Table \@ref(tab:parameters)

RMSE

The closest fit to the observed data was for woody biomass, which had a RMSE of 0.08947263 $kgC/m^2$ and a $R^2$ of 0.9772768 (Supplemental figure \@ref(fig:calibrate))

RMSE
> RMSE
[1] 0.06703855 0.08947263 0.61244709

Process error was approximated as...
> Ql
[1] 0.01040558
> Qw
[1] 0.03424142
> Qs
[1] 0.1499524

## Cv of Q
> Ql/mean(diff(Cl),na.rm=TRUE)
[1] 1.137623
> Qw/mean(diff(Cw),na.rm=TRUE)
[1] 0.1658704

Systematic bias in leaf C (model is ~2x too productive)

Systematic bias in NEE, most likely associated with ...

Overall, while the calibration does suggest room for improvement in the underlying model, it performed well for our target variable (AGB) and appears adequate for the job of testing the multinomial model against real AGB data for a range of different disturbance events.
